---
title: "Chinook salmon harvest"
subtitle: "Exploring the effect of sector on Chinook salmon harvest in the Aleutian Islands"
author: Marina Kochuten
date: last-modified
format: html
editor_options: 
  chunk_output_type: console
---

## Background 

## Data details

## Hypothesis

Does sector have an effect on Chinook salmon harvest? 
- Null: Sector does not have an effect on Chinook salmon harvest
- Alternative: Sector does have an effect on Chinook salmon harvest

## Analysis

#### Setup
```{r}
# Load libraries
library(tidyverse)

# Load data
harvest <- read_csv("Harvest_All_Sectors.csv")
```

#### Data cleaning

The data includes harvest counts for all Alaska regions. For this analysis, I am only interested in the Alaska Peninsula and Aleutian Islands:
```{r}
# Filter harvest data to the Alaska Peninsula and Aleutian Islands
apai_harvest <- harvest |>
  filter(SASAP.Region == "Alaska Peninsula and Aleutian Islands")
```

Next, I want to categorize sector into two groups: subsistence and commercial. Personal use and subsistence catch are hard to divide, as often personal use catch is used as defined by subsistence Because of this, I will group personal use and subsistence into one category (Subsistence). 
```{r}
# Replace all Personal Use with Subsistence
apai_harvest$sector <- str_replace_all(apai_harvest$sector, "Personal Use", "Subsistence")

# Check that Personal Use is no longer in dataset
unique(apai_harvest$sector)
```

I want to compare Chinook salmon against all other species. Let's group all other salmon species into one group, "other". Then, I will be bale to group by species and sector to get total harvest for each species in each sector. Then, I will finish my data cleaning by filtering down to the two sectors of interest, Subsistence and Commercial.
```{r}
chinook_harvest <- apai_harvest |>
  mutate(species = ifelse(species == "chinook", "chinook", "other")) |>
  filter((sector == "Commercial") | (sector == "Subsistence")) |>
  group_by(year, sector, species) |>
  summarise(harvest = sum(harvest, na.rm = TRUE), .groups = "drop")

chinook_grouped <- chinook_harvest |>
  group_by(species, sector) |>
  summarise(species_harvest = sum(harvest, na.rm = TRUE), .groups = "drop") |>
  group_by(sector) |>
  summarize(proportion = species_harvest[species == "chinook"] / sum(species_harvest))
```

#### Visualize the relationship
```{r}
ggplot(chinook_harvest, aes(x = sector, y = harvest, fill = species)) +
  geom_boxplot() +
  facet_wrap(~sector, scale = "free") +
  theme_minimal()
```

```{r}
ggplot(chinook_grouped, aes(x = sector, y = species_harvest, fill = factor(species))) +
  geom_col() +
  facet_wrap(~sector, scale = "free") +
  theme_minimal()
```

#### Hypothesis testing by randomization

**Step 1:** state the null and alternative hypotheses
- Null: Sector does not have an effect on Chinook salmon harvest
- Alternative: Sector does have an effect on Chinook salmon harvest

**Step 2:** calculate the point statistic
- The relevant sample statistic for my hypothesis is a difference in proportions between chinook total harvest / other total harvest in the commercial and subsistence sectors

```{r}
# Find chinook proportion of total harvest in each sector
chinook_grouped <- chinook_harvest |>
  group_by(species, sector) |>
  summarise(species_harvest = sum(harvest, na.rm = TRUE), .groups = "drop") |>
  group_by(sector) |>
  summarize(proportion = species_harvest[species == "chinook"] / sum(species_harvest))

# Calculate difference in proportions between sectors
pe <- chinook_grouped$proportion[2] - chinook_grouped$proportion[1]

pe
```

**Step 3:** quantify the uncertainty
- Use randomization to simulate the distribution of the sample statistic under the null hypothesis.
```{r}
null_dist <- replicate(1000, {
chinook_grouped <- chinook_harvest |>
  mutate(sector = sample(sector, n()),
         species = sample(species, n())) |>
  group_by(species, sector) |>
  summarise(species_harvest = sum(harvest, na.rm = TRUE), .groups = "drop") |>
  group_by(sector) |>
  summarize(proportion = species_harvest[species == "chinook"] / sum(species_harvest))
  
  # Calculate difference in proportions between sectors
  pe <- chinook_prop$proportion[2] - chinook_grouped$proportion[1]
  
  pe
})

ggplot(tibble(null_dist), aes(null_dist)) +
  geom_histogram(bins = 20, color = "cornflowerblue", fill = NA) +
  geom_vline(xintercept = pe, color = "firebrick")
```

**Step 4:** calculate probability of the point estimate under the null
```{r}
sum(abs(null_dist) > abs(pe)) / length(null_dist)
```


