---
title: "Chinook salmon harvest"
subtitle: "Exploring the effect of sector on Chinook salmon harvest in the Aleutian Islands"
author: Marina Kochuten
date: last-modified
format: html
editor_options: 
  chunk_output_type: console
---

## Background 

## Data details

## Hypothesis

Does sector have an effect on Chinook salmon harvest? 
- Null: Sector does not have an effect on Chinook salmon harvest
- Alternative: Sector does have an effect on Chinook salmon harvest

## Analysis

#### Setup
```{r}
# Load libraries
library(tidyverse)

# Load data
harvest <- read_csv("Harvest_All_Sectors.csv")
```

#### Data cleaning

The data includes harvest counts for all Alaska regions. For this analysis, I am only interested in the Alaska Peninsula and Aleutian Islands:
```{r}
# Filter harvest data to the Alaska Peninsula and Aleutian Islands
apai_harvest <- harvest |>
  filter(SASAP.Region == "Alaska Peninsula and Aleutian Islands")
```

Next, I want to categorize sector into two groups: subsistence and commercial. Personal use and subsistence catch are hard to divide, as often personal use catch is used as defined by subsistence Because of this, I will group personal use and subsistence into one category (Subsistence). 
```{r}
# Replace all Personal Use with Subsistence
apai_harvest$sector <- str_replace_all(apai_harvest$sector, "Personal Use", "Subsistence")

# Check that Personal Use is no longer in dataset
unique(apai_harvest$sector)
```

I want to compare Chinook salmon against all other species. Let's group all other salmon species into one group, "other". Then, I will be able to group by species and sector to get total harvest for each species in each sector. Then, I will finish my data cleaning by filtering down to the two sectors of interest, Subsistence and Commercial.
```{r}
chinook_harvest <- apai_harvest |>
  mutate(species = ifelse(species == "chinook", "chinook", "other")) |>
  filter((sector == "Commercial") | (sector == "Subsistence")) |>
  group_by(year, sector, species) |>
  summarize(harvest = sum(harvest, na.rm = TRUE), .groups = "drop")
```

#### Visualize the relationship
```{r}
ggplot(chinook_harvest, aes(x = sector, y = harvest, fill = species)) +
  geom_boxplot() +
  facet_wrap(~sector, scale = "free") +
  scale_fill_brewer("Species", palette = "Pastel1") +
  labs(x = "Number of individuals harvested",
       y = "Fishing Sector",
       title = "Distribution of species harvested by sector") +
  theme_bw()
```

```{r}
ggplot(chinook_harvest, aes(x = sector, y = harvest, fill = factor(species))) +
  geom_col() +
  facet_wrap(~sector, scale = "free") +
  scale_fill_brewer("Species", palette = "Pastel1") +
  labs( x = "Fishing Sector", 
        y = "Number of individuals harvested",  
        title = "Total Chinook Harvest by Sector") +
  theme_minimal()
```

#### Hypothesis testing by randomization

**Step 1:** state the null and alternative hypotheses
- Null: Sector does not have an effect on Chinook salmon harvest
- Alternative: Sector does have an effect on Chinook salmon harvest

**Step 2:** calculate the point statistic
- The relevant sample statistic for my hypothesis is a difference in proportions between chinook total harvest / other total harvest in the commercial and subsistence sectors

```{r}
# Find chinook proportion of total harvest in each sector
chinook_prop <- chinook_harvest |>
  group_by(species, sector) |>
  summarise(species_harvest = sum(harvest, na.rm = TRUE), .groups = "drop") |>
  group_by(sector) |>
  summarize(proportion = species_harvest[species == "chinook"] / sum(species_harvest))

# Calculate difference in proportions between sectors
pe <- chinook_prop$proportion[2] - chinook_prop$proportion[1]
pe
```

**Step 3:** quantify the uncertainty
- Use randomization to simulate the distribution of the sample statistic under the null hypothesis.
```{r}
set.seed(2222)

null_dist <- replicate(1000, {
  chinook_prop <- chinook_harvest |>
    mutate(sector = sample(sector, n())) |>
    group_by(species, sector) |>
    summarise(species_harvest = sum(harvest, na.rm = TRUE), .groups = "drop") |>
    group_by(sector) |>
    summarize(proportion = species_harvest[species == "chinook"] / sum(species_harvest))
  
  # Calculate difference in proportions between sectors
  pe <- chinook_prop$proportion[2] - chinook_prop$proportion[1]
  pe
})

ggplot(tibble(null_dist), aes(null_dist)) +
  geom_histogram(bins = 20, color = "#B8CCE1", fill = NA, linewidth = 1) +
  geom_vline(xintercept = pe, color = "#F0B7B0", linewidth = 1) +
  labs(x = "Null distribution",
       y = "Count") +
  theme_bw()
```

**Step 4:** calculate probability of the point estimate under the null
```{r}
pval <- sum(abs(null_dist) > abs(pe)) / length(null_dist)
print(pval)
```

The p-value 0 is less than the threshold, 0.05. So, I am able to reject the null hypothesis and say that there is a 0% chance that the relationship is due to random chance. I find that sector does have an effect on the proportion of Chinook harvest. 

